Visualization and analysis of probability distributions of large temporal data

### 2022-02-23 report/response.Rmd

> Q18. In Table 3.2, I would suggest separating the m = 1 and m = 2 tables.

Not changed.

> Q19. The presentation of the material on the simulation study on page 48 could be improved. Where is the notation $wpd_{l,s}$ used? Figure 3.5 shows the m = 2 results only. The text alludes to a simulation involving different underlying distributions, but this is not mentioned again.

The simulation design and results corresponding to $m=2$ are presented. Similar design and results for $m=1$, although important, are not included in the paper but in the supplements. Simulation results involving different underlying distributions are also presented in the supplementary paper.

### 2022-02-23 Rmd/03-hakear.Rmd
Catherine Q5 Ch3

Simulations were carried out to explore the behavior of $\wpd$ as $\nx$ and $\nf$ were varied, in order to compare and evaluate different normalization approaches. More detailed simulation results, including an exploration of the size of the proposed test, are contained in the [supplements](https://github.com/Sayani07/paper-hakear/blob/master/paper/paper-supplementary.pdf).

@@

\noindent $m=1$

\noindent Observations were generated from a N(0,1) distribution for $\nx \in \{2, 3, 5, 7, 9, 14, 17, 20, 24, 31, 42, 50\}$, with $\nsub{times} = 500$ observations drawn for each combination of categories. This design corresponds to $\Dnull$. For each of the categories, there were $\nsub{sim}=200$ replications, so that the distribution of $\wpd$ under $\Dnull$ could be observed. Let $\wpd_{\ell, s}$ denote the value of $\wpd$ obtained for the $\ell^{th}$ panel and $s^{th}$ simulation.

\noindent $m=2$

\noindent Similarly, observations were generated from a N(0,1) distribution for each combination of $\nx$ and $\nf$ from $\{2, 3, 5, 7, 14, 20, 31, 50\}$, with $\nsub{times} = 500$ observations drawn for each of the $64$ combinations.

@@

Let $v_{i, t}$ denote the electricity demand for the $i^{th}$ household in time period $t$. The series $\{v_{i, 1},\dots,v_{i,T}\}$ is the linear granularity corresponding to half-hour since the interval of the `tsibble` is 30 minutes. We consider coarser linear granularities like hour, day, week and month from the commonly used Gregorian calendar. From the four linear granularities of hour, day, week, and month, we obtain $N_C = 4\times 3/2 = 6$ cyclic granularities:  "hour_day", "hour_week", "hour_month", "day_ week", "day_month" and "week_month" abbreviated as *hod*, *how*, *hom*, *dow*, *dom* and *wdwnd* respectively. Further, we add cyclic granularity day-type ("wknd wday") to capture weekend and weekday behavior. Thus, seven cyclic granularities are considered to be of interest. The pairs of cyclic granularities ($C_{N_C}$) will have $7\times 6=42$ elements. The set of possible harmonies $H_{N_C}$ from $C_{N_C}$ are chosen by removing clashes using procedures described in @Gupta2021-gravitas. Table \ref{tab:tab-rank-8} shows $14$ harmony pairs that belong to $H_{N_C}$.

@@

Figure \ref{fig:dotplot-8}(c) shows a parallel coordinate plot across different harmonies with harmonies arranged from highest to lowest $wpd$ values averaged over all households. This display is useful for comparing households across harmonies. For example, for the harmony pair (*dow*-*hod*), household id 7 has the greatest value of $\wpd$, while id 1 has the least. From Table \ref{tab:tab-rank-8} it can be seen that the harmony pair (*dow*, *hod*) is important for id 7; however, it has been labeled as an insignificant pair for id 1. The distribution of energy demand for both of these households, with *dow* as the facet and *hod* on the x-axis, may help explain the choice. Figure \ref{fig:gravitas-plot-8} demonstrates that for id 7, the median (black) and quartile deviation (orange) of energy consumption fluctuates for most hours of the day and days of the week, while for id 1, daily patterns are more consistent within weekdays and weekends. As a result, for id 1, it is more appropriate to examine the distributional difference solely across (*dow*, *wdwnd*), which has been rated higher in Table \ref{tab:tab-rank-8}.

@@

<!-- Figure \ref{fig:gravitas-plot-8} is used to show -->
<!-- if this selection and ranking of harmony pairs makes sense for this household. Panel a) of Figure \ref{fig:gravitas-plot-8} shows the distribution of energy demand with weekday/weekend as the x-axis and hour-of-day as the facets and helps to compare the weekend/weekday patterns for different hours of the day. It could be observed that the difference between weekend and weekday is the highest from 15 to 19 hours of the day. Panel b) shows the distribution of energy demand with the variables swapped and helps to compare the daily patterns within weekday and weekend. It could be observed that the daily pattern is similar for weekdays and weekends with a morning and evening peak. However, the difference between morning and evening peaks are higher for weekends. Since $\wpd$ is designed to put more weightage on within-facet differences for $\lambda>0.5$, it makes sense that the pair (hod, wdwnd) has been ranked higher than (wdwnd, hod). Panel c) shows the distribution of energy demand with weekday/weekend as the x-axis and week-of-month as the facets. Although the differences might seem significant at first, with closer inspection it could be seen that the scale of the demand is lower in this case and hence the differences are not large enough to cross the threshold for significance. -->

(ref:gravitas-plot-8) Comparing distribution of energy demand shown for household id 1 (a) and 7 (b) on logarithmic scale across *hod* in x-axis and *dow* in facets through quantile area plots. The value of $\wpd$ in Table 3 suggests that the harmony pair (*dow*, *hod*) is significant for household id 7, but not for id 1. This implies that distributional differences are captured more by this harmony for id 7, which is apparent from this display with more fluctuations across median and 75th percentile for different hours of the day and day of week. For id 1, patterns look similar within different days of weekdays and weekends. Here, the median is represented by the black line, the orange area corresponds to quartile deviation and the green area corresponds to area between $10^{th}$ and $90^{th}$ quantile. The display with *hod* vs *wdwnd* would have shown higher values of wpd for id:1 (as can be verified from Table 3).

>>>>

Simulations were carried out to explore how the behavior of $\wpd$ as $\nx$ and $\nf$ were varied, in order to compare and evaluate different normalization approaches for both $m=1$ and $m=2$. Here the simulation design and results corresponding to $m=2$ are presented. Similar design and results for $m=1$, although important, are not included in the paper but in the [supplements](https://github.com/Sayani07/paper-hakear/blob/master/paper/paper-supplementary.pdf) along with other more detailed simulation results.

@@

<!-- \noindent $m=1$ -->

<!-- \noindent Observations were generated from a N(0,1) distribution for $\nx \in \{2, 3, 5, 7, 9, 14, 17, 20, 24, 31, 42, 50\}$, with $\nsub{times} = 500$ observations drawn for each combination of categories.  Let $\wpd_{\ell, s}$ denote the value of $\wpd$ obtained for the $\ell^{th}$ panel and $s^{th}$ simulation. -->

<!-- \noindent $m=2$ -->

\noindent Observations were generated from a N(0,1) distribution for each combination of $\nx$ and $\nf$ from $\{2, 3, 5, 7, 14, 20, 31, 50\}$, with $\nsub{times} = 500$ observations drawn for each of the $64$ combinations. This design corresponds to $\Dnull$. For each of the categories, there were $\nsub{sim}=200$ replications, so that the distribution of $\wpd$ under $\Dnull$ could be observed.

@@

Let $v_{i, t}$ denote the electricity demand for the $i^{th}$ household in time period $t$. The series $\{v_{i, 1},\dots,v_{i,T}\}$ is the linear granularity corresponding to half-hour since the interval of the `tsibble` is 30 minutes. We consider coarser linear granularities like hour, day, week and month from the commonly used Gregorian calendar. From the four linear granularities of hour, day, week, and month, we obtain $N_C = 4\times 3/2 = 6$ cyclic granularities:  "hour_day", "hour_week", "hour_month", "day_ week", "day_month" and "week_month" abbreviated as *hod*, *how*, *hom*, *dow*, *dom* and *wnwd* respectively. Further, we add cyclic granularity day-type ("wknd wday") to capture weekend and weekday behavior. Thus, seven cyclic granularities are considered to be of interest. The pairs of cyclic granularities ($C_{N_C}$) will have $7\times 6=42$ elements. The set of possible harmonies $H_{N_C}$ from $C_{N_C}$ are chosen by removing clashes using procedures described in @Gupta2021-gravitas. Table \ref{tab:tab-rank-8} shows $14$ harmony pairs that belong to $H_{N_C}$.

@@

Figure \ref{fig:dotplot-8}(c) shows a parallel coordinate plot across different harmonies with harmonies arranged from highest to lowest $wpd$ values averaged over all households. This display is useful for comparing households across harmonies. For example, for the harmony pair (*dow*-*hod*), household id 7 has the greatest value of $\wpd$, while id 1 has the least. From Table \ref{tab:tab-rank-8} it can be seen that the harmony pair (*dow*, *hod*) is important for id 7; however, it has been labeled as an insignificant pair for id 1. The distribution of energy demand for both of these households, with *dow* as the facet and *hod* on the x-axis, may help explain the choice. Figure \ref{fig:gravitas-plot-8} demonstrates that for id 7, the median (black) and quartile deviation (orange) of energy consumption fluctuates for most hours of the day and days of the week, while for id 1, daily patterns are more consistent within weekdays and weekends. As a result, for id 1, it is more appropriate to examine the distributional difference solely across (*dow*, *wnwd*), which has been rated higher in Table \ref{tab:tab-rank-8}.

@@

<!-- Figure \ref{fig:gravitas-plot-8} is used to show -->
<!-- if this selection and ranking of harmony pairs makes sense for this household. Panel a) of Figure \ref{fig:gravitas-plot-8} shows the distribution of energy demand with weekday/weekend as the x-axis and hour-of-day as the facets and helps to compare the weekend/weekday patterns for different hours of the day. It could be observed that the difference between weekend and weekday is the highest from 15 to 19 hours of the day. Panel b) shows the distribution of energy demand with the variables swapped and helps to compare the daily patterns within weekday and weekend. It could be observed that the daily pattern is similar for weekdays and weekends with a morning and evening peak. However, the difference between morning and evening peaks are higher for weekends. Since $\wpd$ is designed to put more weightage on within-facet differences for $\lambda>0.5$, it makes sense that the pair (hod, wnwd) has been ranked higher than (wnwd, hod). Panel c) shows the distribution of energy demand with weekday/weekend as the x-axis and week-of-month as the facets. Although the differences might seem significant at first, with closer inspection it could be seen that the scale of the demand is lower in this case and hence the differences are not large enough to cross the threshold for significance. -->

(ref:gravitas-plot-8) Comparing distribution of energy demand shown for household id 1 (a) and 7 (b) on logarithmic scale across *hod* in x-axis and *dow* in facets through quantile area plots. The value of $\wpd$ in Table 3 suggests that the harmony pair (*dow*, *hod*) is significant for household id 7, but not for id 1. This implies that distributional differences are captured more by this harmony for id 7, which is apparent from this display with more fluctuations across median and 75th percentile for different hours of the day and day of week. For id 1, patterns look similar within different days of weekdays and weekends. Here, the median is represented by the black line, the orange area corresponds to quartile deviation and the green area corresponds to area between $10^{th}$ and $90^{th}$ quantile. The display with *hod* vs *wnwd* would have shown higher values of wpd for id:1 (as can be verified from Table 3).

### 2022-02-23 Rmd/04-gracsr.Rmd
Ch1 Q1 Catherine
Catherine Q5 Ch3

Computationally in R, the data is assumed to be a "tsibble object" [@wang2020tsibble] equipped with an index variable representing inherent ordering from past to present and a key variable that defines observational units over time. The measured variable for an observation is a time-indexed sequence of values. This sequence, however, could be shown in several ways. A shuffle of the raw sequence may represent hourly consumption throughout a day, a week, or a year. Cyclic granularities can be expressed in terms of the index set in the tsibble data structure.

The data object will change when cyclic granularities are computed, as multiple observations will be categorized into levels of the granularity, thus inducing multiple probability distributions. Directly computing Jensen-Shannon distances between the entire probability distributions can be computationally intensive. Thus it is recommended that quantiles are used to characterize the probability distributions. In the final data object, each category of a cyclic granularity corresponds to a list of numbers which is composed of a few quantiles.

@@

There are a few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, we have assumed the time series to be stationary, and hence the distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach cannot detect these dynamic changes. Thirdly, it is possible that for a few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers with such structured missingness. Finally, $\wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers. Moreover, experiments can also be run with non-hierarchy based clustering algorithms to verify if these distances work better with other algorithms.

>>>>

The initial data in R is supposed to be a "tsibble object" [@wang2020tsibble] with an index variable representing inherent ordering from past to present, a key variable that specifies observational units through time, and measured variables. As a result, the measured variable for a key is a sequence of values that is time-indexed. However, this sequence may be shown in a variety of ways. A shuffling of the raw sequence may reflect hourly consumption over the course of a day, a week, or a year.

The data object will change when cyclic granularities are computed, as multiple observations will be categorized into levels of the granularity, thus inducing multiple probability distributions. Directly calculating Jensen-Shannon distances between all probability distributions can be time-consuming. As a result, it is suggested that quantiles be employed to characterize probability distributions. In the final data object, each category of cyclic granularity corresponds to a list of numbers, which is composed of a few quantiles.

@@

There are a few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, the conditional distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach cannot detect these dynamic changes. Thirdly, it is possible that for a few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers with such structured missingness. Finally, $\wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers. Moreover, experiments can also be run with non-hierarchy based clustering algorithms to verify if these distances work better with other algorithms.
